Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data

### 2021-11-25 analysis/paper/paper.Rmd
method gam detal

### Dendrochronological analyses
- GAM for cambial age of urban trees, split like @pretzsch2017 (<=1960, >1960)
-

@@

g(E(DBH_i)) &=  f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i \\

g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\

g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\

@@

where $g()$ is a log-link for $DBH \sim gamma$, and $i$, $j$ and $k$ are indices for observations, species and districts, respectively. 
Above models were tested in combination with a global spatial smooth $f(utmx_i,utmy_i)$ to account for the spatial structure of observations, which reduced auto-correlation of model residuals to considerably (see supplemental information).
**what's the impact of remaining auto correlation? inflated errors, larger coefficients?**
As the spatial extent and coverage varies between temperature and ancillary data, more complex models typically also had fewer total observations, which prevents a direct comparison with information-based model selection criteria, such as Aikake's Information Criterion.
We thus selected the model with the highest explanatory power, and comparatively larger sample size for final analyses.
**further analyses with equal sample sizes (requires data collection / cutting down); but this is first look at potential of approach and initial insights**

>>>>

- models fit using mgcv **citation** in R **citation**

### Dendrochronological analyses

- GAM for cambial age of urban trees, split like @pretzsch2017 (<=1960, >1960)
- lme4, used AR process to account for autocorrelation between observations
- calculataed model averages for all years within year groups

@@

g(E(DBH_i)) &=  f_j(age_i) + f_j(temp_i, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i \\

g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\

g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i, age_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\

@@

where $g()$ is a log-link for $DBH \sim gamma$, and $i$, $j$ and $k$ are indices for observations, species and districts, respectively. 
Models were implemented in `R` [@rcoreteam2021] with `mgcv::bam()` [@wood2017a;@li2020a] for a dramatic increase computational efficiency, allowing for rapid model development/comparison
All models above were also tested with a global spatial smooth $f(utmx_i,utmy_i)$ using Gaussian process smooth [cf. p. 242 in @wood2017] to account for the spatial structure of observations, which reduced auto-correlation of model residuals to considerably (see supplemental information).
**what's the impact of remaining auto correlation? inflated errors, larger coefficients?**

- $f_j$ for temperature interaction with age, implemented as tensor smooth [@wood2006], assuming that temperature impacts vary across trees' lives.
- Reader is referred to detailed implementation in the supplemental material code

As the spatial extent and coverage varies between temperature and ancillary data, more complex models typically also had fewer total observations, which prevents a direct comparison with information-based model selection criteria, such as Aikake's Information Criterion.
We thus selected the model with the highest explanatory power, and comparatively larger sample size for final analyses.
**further analyses with equal sample sizes (requires data collection / cutting down); but this is first look at potential of approach and initial insights**
